---
title: "Process trait data"
author: "E Schlatter"
date: "10/13/2022"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(MasterBayes) #for pedigree functions
library(MCMCglmm) #for running the model
library(MCMCpack) #for inverse wishart distributions
library(bayesplot) #mcmc_areas plot function
library(gridExtra) #multiple plots per figure
source('functions.R')
```


## Fluorescence

### Data processing

Create the following objects we'll use going forward:

$\textbf{fluor}$: Contains blue, green, and red fluorescence values. (Each parent also has a row in this list, to be compatible with relatedness matrix, but its traits and covariates are all NA.)

$\textbf{Ainv}$: the inverse of the relatedness matrix, A. In A, $A_{ij}=$probability of identity by descent of an allele in individuals i and j.

```{r data cleanup}
#fluorescence
fluor <- as.data.frame(read.csv('../traits/fluor_Final_nosize.csv',header=TRUE)) %>%
  rename(green=geren)
fluor <- mutate(fluor,id=as.character(1:nrow(fluor)))

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(fluor,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to fluor for parents
parents <- subset(full_pedig,is.na(dam)) %>%
  transmute(rep=NA,blue=NA,green=NA,red=NA,sire=NA,dam=NA,id=id)
fluor <- rbind(parents,fluor)

fluor$id <- as.factor(fluor$id)
fluor$rep <- as.factor(fluor$rep)
fluor$sire <- as.factor(fluor$sire)
fluor$dam <- as.factor(fluor$dam)

fluor <- fluor %>% 
  rename(animal=id) %>%
  mutate(red_rel = red/(red+blue+green)) %>%
  mutate(green_rel = green/(red+blue+green)) %>%
  mutate(blue_rel = blue/(red+blue+green))

#save(fluor,Ainv,list=c('fluor','Ainv'),file='data/fluor.RData')
```




## Settlement

```{r}
settlement <- as.data.frame(read.csv('../traits/A.millepora_Settlement_2011.csv',header=TRUE)) %>%
  mutate(prop = rec/(rec+lar), id = 1:240) %>%
  rename(dam=mom,sire=dad,timepoint=time)

##### just do the 24-hour timepoint first #######
settlement_24 <- subset(settlement,timepoint=='24')

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_24,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_24 <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_24 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_24),
                                   dimnames = list(NULL,colnames(settlement_24))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_24 <- rbind(parents_add,settlement_24)

#tidy up
settlement_24 <- settlement_24 %>%
  rename(animal=id) %>%
  mutate(across(c(pl,dam,sire,rep,timepoint,animal),factor))

save(settlement_24,Ainv_24,file='data/settlement_24.RData')


##### then the 48-hour timepoint #######
settlement_48 <- subset(settlement,timepoint=='48')

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_48,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_48 <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_24 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_48),
                                   dimnames = list(NULL,colnames(settlement_48))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_48 <- rbind(parents_add,settlement_48)

#tidy up
settlement_48 <- settlement_48 %>%
  rename(animal=id) %>%
  mutate(across(c(pl,dam,sire,rep,timepoint,animal),factor))

save(settlement_48,Ainv_48,file='data/settlement_48.RData')
```


## Lipids

```{r}

lipids <- as.data.frame(read.csv('../traits/Lipid_2011_AC.csv',header=TRUE)) %>%
  rename(dam=mom,sire=dad,timepoint=time) %>%
  mutate(loss=(lipid-egg)/egg)

save(lipids,file='data/lipids.RData')
```

## Protein

```{r}

protein <- as.data.frame(read.csv('../traits/Protein_Orpheus.csv',header=TRUE)) %>%
  rename(dam=mom,sire=dad,egg=wgg,day5=day3) %>% #sarah said day 3 should be day 5
  mutate(loss=(day5-egg)/egg) %>%
  dplyr::select(-day3av)

save(protein,file='data/protein.RData')
```