---
title: "Process trait data"
author: "E Schlatter"
date: "10/13/2022"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(MasterBayes) #for pedigree functions
library(MCMCglmm) #for running the model
library(MCMCpack) #for inverse wishart distributions
library(bayesplot) #mcmc_areas plot function
library(gridExtra) #multiple plots per figure
source('functions.R')
```


## Fluorescence

### Data processing

Create the following objects we'll use going forward:

$\textbf{fluor}$: Contains blue, green, and red fluorescence values. (Each parent also has a row in this list, to be compatible with relatedness matrix, but its traits and covariates are all NA.)

$\textbf{Ainv}$: the inverse of the relatedness matrix, A. In A, $A_{ij}=$probability of identity by descent of an allele in individuals i and j.

```{r data cleanup}
#fluorescence
fluor <- as.data.frame(read.csv('../traits/fluor_Final_nosize.csv',header=TRUE)) %>%
  rename(green=geren)
fluor <- mutate(fluor,id=as.character(1:nrow(fluor)))

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(fluor,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to fluor for parents
parents <- subset(full_pedig,is.na(dam)) %>%
  transmute(rep=NA,blue=NA,green=NA,red=NA,sire=NA,dam=NA,id=id)
fluor <- rbind(parents,fluor)

fluor$id <- as.factor(fluor$id)
fluor$rep <- as.factor(fluor$rep)
fluor$sire <- as.factor(fluor$sire)
fluor$dam <- as.factor(fluor$dam)

fluor <- fluor %>% 
  rename(animal=id) %>%
  mutate(red_rel = red/(red+blue+green)) %>%
  mutate(green_rel = green/(red+blue+green)) %>%
  mutate(blue_rel = blue/(red+blue+green))

#save(fluor,Ainv,list=c('fluor','Ainv'),file='data/fluor.RData')
```




## Settlement

```{r}
settlement <- as.data.frame(read.csv('../traits/A.millepora_Settlement_2011.csv',header=TRUE)) %>%
  mutate(prop = rec/(rec+lar), id = 1:240) %>%
  rename(dam=mom,sire=dad,timepoint=time)

##### just do the 24-hour timepoint first #######
settlement_24 <- subset(settlement,timepoint=='24')

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_24,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_24 <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_24 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_24),
                                   dimnames = list(NULL,colnames(settlement_24))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_24 <- rbind(parents_add,settlement_24)

#tidy up
settlement_24 <- settlement_24 %>%
  rename(animal=id) %>%
  mutate(across(c(pl,dam,sire,rep,timepoint,animal),factor))

save(settlement_24,Ainv_24,file='data/settlement_24.RData')


##### then the 48-hour timepoint #######
settlement_48 <- subset(settlement,timepoint=='48')

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_48,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_48 <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_24 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_48),
                                   dimnames = list(NULL,colnames(settlement_48))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_48 <- rbind(parents_add,settlement_48)

#tidy up
settlement_48 <- settlement_48 %>%
  rename(animal=id) %>%
  mutate(across(c(pl,dam,sire,rep,timepoint,animal),factor))

save(settlement_48,Ainv_48,file='data/settlement_48.RData')
```

### Binomial -- 24 hours
```{r}
# Process settlement data for use in a binomial model
# one row per larva. New phenotype, settled, is 0 for larvae and 1 for recruits.
# This is messy, sorry! Maybe I'll clean it up sometime.
load('data/settlement_24.RData')

settlement_24 <- settlement_24 %>%
  dplyr::select(animal,dam,sire,pl,rep,rec,lar) %>%
  mutate(clutch = paste(dam,sire,pl,rep,sep='.'))
settlement_24$clutch[1:10] <- NA

#add a's and b's to distinguish clones' clutches from each other
a_inds <- which(settlement_24$animal %in% c(25:36,97:108))
b_inds <- which(settlement_24$animal %in% c(145:156,193:204))
settlement_24$clutch[a_inds] <- paste(settlement_24$clutch[a_inds],'a',sep='')
settlement_24$clutch[b_inds] <- paste(settlement_24$clutch[b_inds],'b',sep='')

settlement_24_binom = data.frame(dam = settlement_24$dam[1:10],
                                 sire = settlement_24$sire[1:10], clutch = settlement_24$clutch[1:10],
                                 settled = NA)
                       
for(cl in 11:nrow(settlement_24)){ #for each clutch
  if(settlement_24$lar[cl]!=0){
    for(larvs in 1:(settlement_24$lar[cl])){ #add as many rows as there are larvae in the clutch, with settled=0
      new_row <- data.frame(dam = settlement_24$dam[cl],
                            sire = settlement_24$sire[cl], clutch = settlement_24$clutch[cl],
                            settled = 0)
      settlement_24_binom <- rbind(settlement_24_binom,new_row)
      }
  }
  if(settlement_24$rec[cl]!=0){
    for(recruits in 1:settlement_24$rec[cl]){ #add as many rows as there are recruits in the clutch, with settled=1
      new_row <- data.frame(dam = settlement_24$dam[cl],
                          sire = settlement_24$sire[cl], clutch = settlement_24$clutch[cl],
                          settled = 1)
    settlement_24_binom <- rbind(settlement_24_binom,new_row)
    }
  }
}
#sum(settlement_24$lar,na.rm=TRUE)+sum(settlement_24$rec,na.rm=TRUE)+10
#check that we did it right: this matches the number of rows in the new dataframe

settlement_24_binom <- filter(settlement_24_binom,!is.na(sire)) #get rid of parent rows
settlement_24_binom$id <- as.factor(1:nrow(settlement_24_binom)) #add id column

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_24_binom,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_24_binom <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_24 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_24_binom),
                                   dimnames = list(NULL,colnames(settlement_24_binom))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_24_binom <- rbind(parents_add,settlement_24_binom) %>%
  rename(animal=id)


save(settlement_24_binom,Ainv_24_binom,file='data/settlement_24_binom.RData')
```
### Binomial -- 48 hours
```{r}
# Process settlement data for use in a binomial model
# one row per larva. New phenotype, settled, is 0 for larvae and 1 for recruits.
load('data/settlement_48.RData')

settlement_48 <- settlement_48 %>%
  dplyr::select(animal,dam,sire,pl,rep,rec,lar) %>%
  mutate(clutch = paste(dam,sire,pl,rep,sep='.'))
settlement_48$clutch[1:10] <- NA

#add a's and b's to distinguish clones' clutches from each other
a_inds <- which(settlement_48$animal %in% c(37:48,109:120))
b_inds <- which(settlement_48$animal %in% c(157:168,205:216))
settlement_48$clutch[a_inds] <- paste(settlement_48$clutch[a_inds],'a',sep='')
settlement_48$clutch[b_inds] <- paste(settlement_48$clutch[b_inds],'b',sep='')

settlement_48_binom = data.frame(dam = settlement_48$dam[1:10],
                                 sire = settlement_48$sire[1:10], clutch = settlement_48$clutch[1:10],
                                 settled = NA)
                       
for(cl in 11:nrow(settlement_48)){ #for each clutch
  if(settlement_48$lar[cl]!=0){
    for(larvs in 1:(settlement_48$lar[cl])){ #add as many rows as there are larvae in the clutch, with settled=0
      new_row <- data.frame(dam = settlement_48$dam[cl],
                            sire = settlement_48$sire[cl], clutch = settlement_48$clutch[cl],
                            settled = 0)
      settlement_48_binom <- rbind(settlement_48_binom,new_row)
      }
  }
  if(settlement_48$rec[cl]!=0){
    for(recruits in 1:settlement_48$rec[cl]){ #add as many rows as there are recruits in the clutch, with settled=1
      new_row <- data.frame(dam = settlement_48$dam[cl],
                          sire = settlement_48$sire[cl], clutch = settlement_48$clutch[cl],
                          settled = 1)
    settlement_48_binom <- rbind(settlement_48_binom,new_row)
    }
  }
}
#sum(settlement_48$lar,na.rm=TRUE)+sum(settlement_48$rec,na.rm=TRUE)+10
#check that we did it right: this matches the number of rows in the new dataframe

settlement_48_binom <- filter(settlement_48_binom,!is.na(sire)) #get rid of parent rows
settlement_48_binom$id <- as.factor(1:nrow(settlement_48_binom)) #add id column

#pedigree and Ainv -- we'll need this later for heritability analysis
pedigree <- dplyr::select(settlement_48_binom,id,dam,sire)

#pedigree utility functions (from MasterBayes) to insert parents and correctly order, and create Ainv
full_pedig <- pedigree %>%
  insertPed() %>%
  orderPed()
Ainv_48_binom <- MCMCglmm::inverseA(full_pedig)$Ainv

#add empty rows to settlement_48 for parents
parents_add = as.data.frame(matrix(nrow = 10, ncol = ncol(settlement_48_binom),
                                   dimnames = list(NULL,colnames(settlement_48_binom))
                                   ))
parents_add$id <- subset(full_pedig,is.na(dam))$id
parents_add$dam <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ') #If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).

settlement_48_binom <- rbind(parents_add,settlement_48_binom) %>%
  rename(animal=id)

save(settlement_48_binom,Ainv_48_binom,file='data/settlement_48_binom.RData')
```


## Lipids

```{r}

lipids <- as.data.frame(read.csv('../traits/Lipid_2011_AC.csv',header=TRUE)) %>%
  rename(dam=mom,sire=dad,timepoint=time) %>%
  mutate(loss=(lipid-egg)/egg)

save(lipids,file='data/lipids.RData')
```

## Protein

```{r}

protein <- as.data.frame(read.csv('../traits/Protein_Orpheus.csv',header=TRUE)) %>%
  rename(dam=mom,sire=dad,egg=wgg,day5=day3) %>% #sarah said day 3 should be day 5
  mutate(loss=(day5-egg)/egg) %>%
  dplyr::select(-day3av)

save(protein,file='data/protein.RData')
```