---
title: "Run heritability analyses"
author: "E Schlatter"
date: "10/13/2022"
output: html_document
---

```{r message=FALSE, warning=FALSE}
setwd("C:/Users/eschlatter/Dropbox/coral_h2")
library(tidyverse)
library(MasterBayes) #for pedigree functions
library(MCMCglmm) #for running the model
library(MCMCpack) #for inverse wishart distributions
library(bayesplot) #mcmc_areas plot function
library(gridExtra) #multiple plots per figure
source('analysis/functions.R')
```

Let's stick to the animal model for now, since it sounds like that's what we'll use for gene expression.

## Fluorescence
```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/data/fluor.RData')
```

### Red

#### red~animal+dam+rep
Makes sense not to try to estimate sire separately from animal, because there's not much replication within sires.

#####5. Inverse Wishart priors w/even division of variance among components
```{r,eval=FALSE}
##First try at estimation using this model. Use priors based on even division of variance components.
nu = 0.001
p_var = var(fluor$red_rel,na.rm=TRUE) #total phenotypic variance in relative green fluor
prior_mode = p_var/4 #prior mode (4 variance components)
V = ((nu+2)/nu)*prior_mode #prior parameter V (inverse Wishart) for additive genetic effect

prior5_red = list(R=list(V=V,nu=nu),
                    G=list(G1=list(V=V,nu=nu),G2=list(V=V,nu=nu),
                           G3=list(V=V,nu=nu)))

red_fluor_5 <- MCMCglmm(red_rel~1,
                   random=~animal+dam+rep,
                   prior=prior5_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=100000,burnin=1000,thin=10,verbose=FALSE)

save(red_fluor_5,prior5_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_5.RData')
```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_5.RData')

summary(red_fluor_5)
plot(red_fluor_5)

traces <- red_fluor_5$VCV
dam_trace <- data.frame(time=1:9900,dam=as.data.frame(traces[,2]))
plot(dam_trace,type='l',ylim=c(0,5))
#x <- dist(dam_trace$var1)


herit <- traces[,1]/rowSums(traces)
plot(herit)
effectiveSize(herit)
HPDinterval(herit)


Herit <- (traces[,1]+traces[,2])/rowSums(traces)
plot(Herit)
effectiveSize(Herit)
HPDinterval(Herit)
```

#####6. Same thing as 5, run longer
```{r,eval=FALSE}
##Try again, just run for longer
nu = 0.001
p_var = var(fluor$red_rel,na.rm=TRUE) #total phenotypic variance in relative green fluor
prior_mode = p_var/4 #prior mode (4 variance components)
V = ((nu+2)/nu)*prior_mode #prior parameter V (inverse Wishart) for additive genetic effect

prior6_red = list(R=list(V=V,nu=nu),
                    G=list(G1=list(V=V,nu=nu),G2=list(V=V,nu=nu),
                           G3=list(V=V,nu=nu)))

red_fluor_6 <- MCMCglmm(red_rel~1,
                   random=~animal+dam+rep,
                   prior=prior6_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

save(red_fluor_6,prior6_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_6.RData')
```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_6.RData')

model <- red_fluor_6
summary(model)
plot(model)

traces <- model$VCV
dam_trace <- data.frame(time=1:9900,dam=as.data.frame(traces[,2]))
plot(dam_trace,type='l',ylim=c(0,5))
#x <- dist(dam_trace$var1)


herit <- traces[,1]/rowSums(traces)
plot(herit)
effectiveSize(herit)
HPDinterval(herit)


Herit <- (traces[,1]+traces[,2])/rowSums(traces)
plot(Herit)
effectiveSize(Herit)
HPDinterval(Herit)
```

#####7. Parameter expansion.
This does a better job on the dam component -- ridiculously large values are a couple orders of magnitude smaller (like 6 instead of 400)
```{r,eval=FALSE}
##Try with parameter expansion
prior7_red = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G3=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

red_fluor_7 <- MCMCglmm(red_rel~1,
                   random=~animal+dam+rep,
                   prior=prior7_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

save(red_fluor_7,prior7_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_7.RData')
```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_7.RData')

model <- red_fluor_7
summary(model)
plot(model)

traces <- model$VCV
dam_trace <- data.frame(time=1:9900,dam=as.data.frame(traces[,2]))
plot(dam_trace,type='l',ylim=c(0,5))
#might just need to run a lot longer? Seems ridiculous. But the high peaks seem to get lower as iterations increase. Longer burnin?

herit <- traces[,1]/rowSums(traces)
plot(herit)
effectiveSize(herit)
HPDinterval(herit)

Herit <- (traces[,1]+traces[,2])/rowSums(traces)
plot(Herit)
effectiveSize(Herit)
HPDinterval(Herit)
```

#####8. Dam as a fixed effect.
The difference between a fixed and a random effect in this type of model is that, for a random effect, you estimate the variance (among levels). For a fixed effect, the variance is set (usually to something large). With only two levels for dam, we don't have enough data to estimate the variance among dams in a meaningful way. So we should treat it as a fixed effect.

```{r,eval=FALSE}
##Try with parameter expansion

#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10).
prior8_red = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

#If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).
fluor$dam <- as.character(fluor$dam)
fluor$dam[1:10] <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ')
fluor$dam <- as.factor(fluor$dam)



red_fluor_8 <- MCMCglmm(red_rel~1+dam,
                   random=~animal+rep,
                   prior=prior8_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=100000,burnin=1000,thin=10,verbose=TRUE)

save(red_fluor_8,prior8_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_8.RData')
```

```{r,eval=FALSE}
##Try with parameter expansion
prior9_red = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

red_fluor_9 <- MCMCglmm(red_rel~dam-1,
                   random=~animal+rep,
                   prior=prior9_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=100000,burnin=1000,thin=10,verbose=TRUE)

save(red_fluor_9,prior9_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_9.RData')

```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_9.RData')

#by including dam as a fixed effect, we artificially inflate the estimate of heritability if it's calculated in the traditional way (VA divided by the sum of estimates of all variance components). To get an estimate that means what we expect heritability to mean, we can instead divide by the observed phenotypic variance of the data.
#[see: Wilson 2008, Journal of Evolutionary Biology: Why h^2 does not always equal VA/VP]
#if we do this, we should check that adding dam as a fixed effect this way doesn't change the estimate of VA too much.
plot(red_fluor_9)
summary(red_fluor_9)
vp_observed <- var(fluor$red_rel,na.rm=TRUE)
herit_obs <- red_fluor_9$VCV[,1]/vp_observed
plot(herit_obs)
summary(herit_obs)
```


```{r,eval=FALSE}
#try non-parameter-expanded
nu = 0.001
p_var = var(fluor$red_rel,na.rm=TRUE) #total phenotypic variance in relative red fluor
prior_mode = p_var/3 #prior mode (3 variance components)
V = ((nu+2)/nu)*prior_mode #prior parameter V (inverse Wishart) for additive genetic effect

prior10_red = list(R=list(V=V,nu=nu),
                    G=list(G1=list(V=V,nu=nu),G2=list(V=V,nu=nu)))

red_fluor_10 <- MCMCglmm(red_rel~dam-1,
                   random=~animal+rep,
                   prior=prior10_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=100000,burnin=1000,thin=10,verbose=TRUE)

save(red_fluor_10,prior10_red,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_10.RData')

```


Not much difference between the parameter-expanded priors and the ones centered on a fraction of the observed phenotypic variance. In this case, that value is probably too high, because the fixed effect (dam) is soaking up a lot of the variance. So, a priori, maybe better to use the parameter expansion.
```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_9.RData')
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor_10.RData')
plot(red_fluor_9)
plot(red_fluor_10)
```


### Blue

#### blue~animal+dam+rep

#####1. Parameter expansion iter=10^6
```{r,eval=FALSE}
##Try with parameter expansion
prior1_blue = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G3=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

blue_fluor_1 <- MCMCglmm(blue_rel~1,
                   random=~animal+dam+rep,
                   prior=prior1_blue,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

save(blue_fluor_1,prior1_blue,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/blue_fluor_1.RData')
```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/blue_fluor_1.RData')

model <- blue_fluor_1
summary(model)
plot(model)

traces <- model$VCV
dam_trace <- data.frame(time=1:9900,dam=as.data.frame(traces[,2]))
plot(dam_trace,type='l',ylim=c(0,5))

herit <- traces[,1]/rowSums(traces)
plot(herit)
effectiveSize(herit)
HPDinterval(herit)

Herit <- (traces[,1]+traces[,2])/rowSums(traces)
plot(Herit)
effectiveSize(Herit)
HPDinterval(Herit)
```
### Green

#### green~animal+dam+rep
(What about dam as a fixed effect? Would that make sense?)
#####1. Parameter expansion iter=10^6
```{r,eval=FALSE}
##Try with parameter expansion
prior1_green = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G3=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

green_fluor_1 <- MCMCglmm(green_rel~1,
                   random=~animal+dam+rep,
                   prior=prior1_blue,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

save(green_fluor_1,prior1_green,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/green_fluor_1.RData')
```

```{r}
load('C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/green_fluor_1.RData')

model <- green_fluor_1
summary(model)
plot(model)

traces <- model$VCV
dam_trace <- data.frame(time=1:9900,dam=as.data.frame(traces[,2]))
plot(dam_trace,type='l',ylim=c(0,5))

herit <- traces[,1]/rowSums(traces)
plot(herit)
effectiveSize(herit)
HPDinterval(herit)

Herit <- (traces[,1]+traces[,2])/rowSums(traces)
plot(Herit)
effectiveSize(Herit)
HPDinterval(Herit)
```