---
title: "A. millepora larval trait heritability"
author: "E Schlatter"
date: "9/26/2023"
output:
  html_document:
    toc: true
    toc_depth: 4

---
## Intro
```{r message=FALSE, warning=FALSE, echo=FALSE}
#setwd("C:/Users/eschlatter/Dropbox/coral_h2")
library(tidyverse)
library(MasterBayes) #for pedigree functions
library(MCMCglmm) #for running the model
library(MCMCpack) #for inverse wishart distributions
library(bayesplot) #mcmc_areas plot function
library(gridExtra) #multiple plots per figure
library(knitr)
source('functions.R')
```


## Fluorescence
```{r, echo=FALSE}
load('data/fluor.RData')
```
The model I've used is **color ~ dam + rep + breeding value + residual.**

Each color is treated separately, for now.

I'm using the relative fluorescence of each color -- e.g., red / (red+blue+green).

I haven't transformed the data in any way. In Sarah's dissertation, she used arcsin(square root(relative fluorescence)).

**Animal Model**

This is an animal model: the "breeding value" term is essentially the contribution of additive genetic effect. (Breeding value is modeled as a normally-distributed random variable, with covariances among individuals according to their relatedness. The coefficient of this random variable is VA.) I've stuck to the animal model for now, since it sounds like that's what we'll use for gene expression.

**Sire effect not estimated separately from VA**

Sire isn't included as a predictor variable here. I think it makes sense not to try to estimate a sire effect separately from breeding value, since there's not much replication within sires. (That is: for each larva, we're already getting an estimate of the contribution of additive genetic effect to their phenotype. Trying to estimate the contribution of paternal ID on top of that -- i.e., how much sharing a father contributes to the similarity between two larvae, over and above the effect from the 1/4 additional relatedness -- is a lot to ask, given the number of offspring per sire.)

**Dam as a fixed effect**

I've modeled dam as a fixed effect. The only difference between a fixed and a random effect in a Bayesian model is that, for a random effect, you estimate the variance (among levels). For a fixed effect, the variance is set (usually to something large). With only two levels for dam, we don't have enough data to estimate the variance among dams in a meaningful way. So we should treat it as a fixed effect.

This introduces a problem when estimating heritability. By including dam as a fixed effect, we artificially inflate the estimate of heritability as it's usually calculated (VA divided by the sum of estimates of all variance components). Since there's no variance component representing the -- substantial! -- variation due to dam, the denominator is artificially small, and h2 is artificially large. [see: Wilson 2008, Journal of Evolutionary Biology: Why h^2 does not always equal VA/VP]

We can address this by estimating the variance due to the fixed effect, and including that in the denominator. I've followed de Villemereuil et al 2018 (eqn 7), and used some code from de Villemereuil's MCMCglmm animal model tutorial (Version 2, 2023, p.26). [Note: this is an improvement over my first try, just using the phenotypic variance as the denominator of h2. It gives quantitatively different results -- h2 of red fluorescence, e.g., is about 0.15 instead of 0.42. But the credible intervals are tighter -- as expected, since, in any given MCMC iteration, if more of the variance is attributed to fixed effects, less is attributed to random effects.]

Alternatively, we could include dam as a random effect, so that its associated variance is included in the denominator when estimating heritability. But, although the model *will* give us an answer when we ask it to estimate the variance attributed to dam, it's 1) not well-behaved (some iterations of the MCMC algorithm give preposterously high values, like 1000), and 2) isn't really meaningful, because it's estimated from only two data points (dam A and dam C).

**Influence of priors**

The prior I'm using in the results shown here is parameter expanded for the non-residual components (rep and animal), and uses the default inverse Wishart for residual (MCMCglmm can't do parameter expansion for the residual).

It's not shown here, but I've run these models with several different priors. They all have good mixing properties, and the estimates are essentially the same. I take that to mean there's enough signal in the data to overcome the influence of the prior we give it -- a good sign!

### Summary

**Red fluorescence**

```{r echo=FALSE, message=FALSE}
load('models/red_fluor.RData')
red_est <- fn_get_estimates(red_fluor,herit_red_fluor)
kable(red_est)
```

**Blue fluorescence**

```{r echo=FALSE, message=FALSE}
load('models/blue_fluor.RData')
# vp_blue <- var(fluor$blue_rel,na.rm=TRUE)
# herit_blue <- blue_fluor$VCV[,1]/vp_blue
# blue_est_old <- fn_get_estimates(blue_fluor,herit_blue)
# kable(blue_est_old)
blue_est <- fn_get_estimates(blue_fluor,herit_blue_fluor)
kable(blue_est)
```

**Green fluorescence**

```{r echo=FALSE, message=FALSE}
load('models/green_fluor.RData')
# vp_green <- var(fluor$green_rel,na.rm=TRUE)
# herit_green <- green_fluor$VCV[,1]/vp_green
# green_est_old <- fn_get_estimates(green_fluor,herit_green)
# kable(green_est_old)
green_est <- fn_get_estimates(green_fluor,herit_green_fluor)
kable(green_est)
```

Detailed output of each of the three models below.

### Red
```{r,eval=FALSE, echo=FALSE}
#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_red = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

#If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).
fluor$dam <- as.character(fluor$dam)
fluor$dam[1:10] <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ')
fluor$dam <- as.factor(fluor$dam)

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
red_fluor <- MCMCglmm(red_rel~dam-1,
                   random=~animal+rep,
                   prior=prior_red,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

## this is the better way to calculate h2 with fixed effects.
# Rather than just dividing by the observed phenotypic variance, use the variance of the predicted values from the fixed effects as VF, and let VP = VF+VA+VE be the denominator.
# (See de Villemereuil 2023)
model <- red_fluor
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X) #multiply to get predicted phenotype, just based on dam, and take the variance of that over all larvae (once for each MCMC iteration)
herit_red_fluor <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(red_fluor,prior_red,herit_red_fluor,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/red_fluor.RData')
```

**MCMCglmm output:**

```{r echo=FALSE, message=FALSE, warning=FALSE}
load('models/red_fluor.RData')
model <- red_fluor

summary(model)
plot(model)
```

**Priors and posteriors:**

```{r echo=FALSE, message=FALSE, warning=FALSE}
fn_plot_all_priorandpost(model,prior_red,xlim=0.005)
```

**Heritability:**

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(herit_red_fluor)
plot(herit_red_fluor)
```

### Blue

```{r,eval=FALSE, echo=FALSE}
#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_blue = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

#If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).
fluor$dam <- as.character(fluor$dam)
fluor$dam[1:10] <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ')
fluor$dam <- as.factor(fluor$dam)

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
blue_fluor <- MCMCglmm(blue_rel~dam-1,
                   random=~animal+rep,
                   prior=prior_blue,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- blue_fluor
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_blue_fluor <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(blue_fluor,prior_blue,herit_blue_fluor,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/blue_fluor.RData')
```

**MCMCglmm output:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
load('models/blue_fluor.RData')
model <- blue_fluor

summary(model)
plot(model)
```

**Priors and posteriors:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
fn_plot_all_priorandpost(model,prior_blue,xlim=0.005)
```

**Heritability:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(herit_blue_fluor)
plot(herit_blue_fluor)
```

### Green

```{r,eval=FALSE, echo=FALSE}
#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_green = list(R=list(V=1,nu=0.002),
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1),
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1)))

#If we treat dam as fixed, then MCMCglmm needs a dam assigned to each individual in the dataset, even the parents. We'll add dummy variables in for them. This is ok, since the parents' rows don't get used anyway (they don't have phenotype data).
fluor$dam <- as.character(fluor$dam)
fluor$dam[1:10] <- c('AA','BB','CC','DD','EE','FF','GG','HH','II','JJ')
fluor$dam <- as.factor(fluor$dam)

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
green_fluor <- MCMCglmm(green_rel~dam-1,
                   random=~animal+rep,
                   prior=prior_green,
                   ginverse=list(animal=Ainv),
                   data=fluor,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- green_fluor
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_green_fluor <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(green_fluor,prior_green,herit_green_fluor,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/green_fluor.RData')
```

**MCMCglmm output:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
load('models/green_fluor.RData')
model <- green_fluor

summary(model)
plot(model)
```

**Priors and posteriors:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
fn_plot_all_priorandpost(model,prior_green,xlim=0.005)
```

**Heritability:**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(herit_green_fluor)
plot(herit_green_fluor)
```

## Settlement

```{r,echo=FALSE}
load('data/settlement.RData')
```

So far, I'm treating the 24-hour and 48-hour timepoints separately.

```{r, echo=FALSE, message=FALSE, fig.dim=c(3,3)}
hist(settlement_24$prop)
hist(settlement_48$prop)

hist(asin(sqrt(settlement_24$prop)))
hist(asin(sqrt(settlement_48$prop)))
```

These aren't normally-distributed data. They're proportions that vary between 0 and 1. The arcsin(sqrt) transformation makes them look a little more normal. But the estimation of variance components on the nonlinearly-transformed data doesn't translate in a straightforward way back to the original (what de Villemereuil et al 2018 calls the latent scale to the data scale). I think the solution to this is a GLMM (see below).

But first, let's try estimating heritability from the untransformed and the transformed data, and see if there's any difference.

- The model I'm using is **settlement proportion ~ dam + breeding value + residual**

- **??** I didn't include rep as a fixed effect. e.g., parents A and P have 6 datapoints at time 24: rep 1 plate A, rep 1 plate E, rep 1 plate F, rep 2 plate L, rep 2 plate O, rep 2 plate R. I'm assuming that 2 plates from the same rep are no more likely to be similar than plates from opposite reps. Is this right? Otherwise we could include rep as a fixed effect w/2 levels (1 or 2).

### Summary

In this section are tables of variance component and heritability estimates, and traces and posterior plots of heritability. More details on each of the model outputs are in the sections that follow.

#### Gaussian model: proportion of recruits

At 24 hours, the posterior credible interval for heritability is very wide. In both cases, the lower boundary of the 95% interval is close to zero. (Closer with the transformed data, which also has a slightly smaller median.) The distribution is sort of a plateau, rather than a nice unimodal shape. I take this to mean there's probably a weak signal, and we don't have enough data to resolve it fully.

**Untransformed data -- 24 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle24.RData')
settle24_est <- fn_get_estimates(settle24,herit_settle24)
kable(settle24_est)
plot(herit_settle24)
```

**Transformed data -- 24 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle24_transform.RData')
settle24_transform_est <- fn_get_estimates(settle24_transform,herit_settle24_transform)
kable(settle24_transform_est)
plot(herit_settle24_transform)
```

At 48 hours, the posterior distributions for heritability are narrower, and definitely different from zero. (Again, the transformed data give slightly smaller estimates.)

The data transformation doesn't make much of a difference in terms of results. Of the two options, I'd go with the untransformed data, because the interpretation is simpler: heritability of the proportion of offspring that recruit, vs heritability of the arcsin(sqrt()) of the proportion of offspring that recruit.

Something that bothers me a little, though, is: whose trait is this? This analysis essentially treats each plate as an individual, and its trait is the proportion of recruits. Biologically speaking, that's not quite right. I'm not sure if it's a problem, but it makes me a little uncomfortable. So I think actually I'd advocate for a third option: a binomial model, described below, which has a simpler interpretation and avoids this problem.

**Untransformed data -- 48 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle48.RData')
settle48_est <- fn_get_estimates(settle48,herit_settle48)
kable(settle48_est)
plot(herit_settle48)
```

**Transformed data -- 48 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle48_transform.RData')
settle48_transform_est <- fn_get_estimates(settle48_transform,herit_settle48_transform)
kable(settle48_transform_est)
plot(herit_settle48_transform)
```

#### Binomial model: recruitment-readiness

In the original dataset, each row is one plate (at either 24 or 48 hours), where Sarah observed $n$ larvae and $m$ recruits. There are 6 plates for each pair of parents at each timepoint. For this analysis, I've reorganized the data so that, instead, each individual offspring (recruit or larva) is its own observation. So for each plate, there are $n+m$ rows in the reorganized dataset: around 3000 individuals at each of the 2 timepoints.

The response variable here isn't the proportion of a plate that have recruited, it's a binary variable (called "settled") that applies to each individual offspring: 0 for larvae and 1 for recruits. We can use a binomial GLMM to look at its heritability.

The GLMM assumes that the binary response variable ("settled") is a function of a latent trait that's normally distributed. In this context, that's saying that whether a larva has settled or not is determined by many genes of small effect. We can interpret this latent trait as meaning something like settlement-readiness. The GLMM assumes each individual has a value for this underlying trait, and there's a threshold: individuals with a value above the threshold are recruits, and those below are larvae.

It's straightforward to calculate heritability in terms of the latent trait, and this is often what's done in the literature. In our case, I think it's a reasonable choice; it's a slight reframing of the question to ask about the heritability of settlement-readiness (at 24 or 48 hours) instead of the heritability of the proportion of offspring settled.

I've included plate as a random effect, to control for the similarity in environment experienced by offspring in the same plate. (I called it `clutch`, because `plate` is already used in this dataset, and the same plate name appears for more than one set of parents.)

I think including clutch means we can leave out dam as a fixed effect. Accounting for the variance among clutches includes the variance among dams. (Right?) I think this is good enough, since we don't specifically need to know the effect of the dams. And calculating heritability in a model with fixed effects is more complicated when it's a GLMM. So I've left it out for now.

The model is **settled ~ breeding value + clutch + residual**.

Qualitatively, the result is the same as for the proportion data: at 24 hours, the 95% posterior credible interval for heritability comes very close to zero (3e-4), while at 48 hours, it's quite distinct (0.12). But the estimates from the binomial model are quite a bit lower than from the (untransformed) proportion data: median of 0.17 vs 0.47 at 24 hours, and median of 0.37 vs 0.66 at 48 hours.

[The residual variance is fixed at 1. That's because the nature of the binomial distribution means we can only estimate the relative sizes of the variance components. That's not a problem since, for heritability, relative magnitudes of variance components are all we care about anyway.]

The GLMM doesn't mix very well yet. It's good enough that I'm pretty confident in the results, but I'd want it to be better for publication. I'm still looking into methods to improve it (trying different priors, etc).

**Binomial model -- 24 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle24_binom.RData')
settle24_binom_est <- fn_get_estimates(settle24_binom,herit_settle24_binom)
kable(settle24_binom_est)
plot(herit_settle24_binom)
```

**Binomial model -- 48 hours**

```{r,echo=FALSE, message=FALSE,fig.dim=c(4.5,3)}
load('models/settle48_binom.RData')
settle48_binom_est <- fn_get_estimates(settle48_binom,herit_settle48_binom)
kable(settle48_binom_est)
plot(herit_settle48_binom)
```

### Untransformed data

#### 24 hours
```{r,eval=FALSE, echo=FALSE}
load('data/settlement_24.RData')

#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_settle24 = list(R=list(V=1,nu=0.002), #R: residual effect
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) #G1: breeding value

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
settle24 <- MCMCglmm(prop~dam-1,
                   random=~animal,
                   prior=prior_settle24,
                   ginverse=list(animal=Ainv_24),
                   data=settlement_24,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle24
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_settle24 <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(settle24,prior_settle24,herit_settle24,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle24.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle24.RData')
model <- settle24
prior <- prior_settle24
herit <- herit_settle24

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
fn_plot_all_priorandpost(model,prior,xlim=0.1)
plot(herit)
```


#### 48 hours
```{r,eval=FALSE, echo=FALSE}
load('data/settlement_48.RData')

#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_settle48 = list(R=list(V=1,nu=0.002), #R: residual effect
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) #G1: breeding value

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
settle48 <- MCMCglmm(prop~dam-1,
                   random=~animal,
                   prior=prior_settle48,
                   ginverse=list(animal=Ainv_48),
                   data=settlement_48,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle48
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_settle48 <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(settle48,prior_settle48,herit_settle48,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle48.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle48.RData')
model <- settle48
prior <- prior_settle48
herit <- herit_settle48

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
fn_plot_all_priorandpost(model,prior,xlim=0.1)
plot(herit)
```


### Transformed data (arcsin(sqrt))

#### 24 hours
```{r,eval=FALSE, echo=FALSE}
load('data/settlement_24.RData')
settlement_24 <- mutate(settlement_24,prop_transform = asin(sqrt(prop)))

#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_settle24_transform = list(R=list(V=1,nu=0.002), #R: residual effect
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) #G1: breeding value

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
settle24_transform <- MCMCglmm(prop_transform~dam-1,
                   random=~animal,
                   prior=prior_settle24_transform,
                   ginverse=list(animal=Ainv_24),
                   data=settlement_24,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle24_transform
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_settle24_transform <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(settle24_transform,prior_settle24_transform,herit_settle24_transform,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle24_transform.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle24_transform.RData')
model <- settle24_transform
prior <- prior_settle24_transform
herit <- herit_settle24_transform

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
fn_plot_all_priorandpost(model,prior,xlim=0.1)
plot(herit)
```

#### 48 hours

```{r,eval=FALSE, echo=FALSE}
load('data/settlement_48.RData')
settlement_48 <- mutate(settlement_48,prop_transform = asin(sqrt(prop)))

#The default fixed effect prior has a zero mean vector and a diagonal variance matrix with large variances (1e+10). We'll leave that alone, and specify the random and residual effect priors as follows:
prior_settle48_transform = list(R=list(V=1,nu=0.002), #R: residual effect
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) #G1: breeding value

#fixed effects written as dam-1 means we don't specify a global intercept; each level of dam (A and C) has its own. 
settle48_transform <- MCMCglmm(prop_transform~dam-1,
                   random=~animal,
                   prior=prior_settle48_transform,
                   ginverse=list(animal=Ainv_48),
                   data=settlement_48,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle48_transform
X <- model[['X']] #design matrix relating observations (larvae) to fixed effects (dams)
VF <- apply(model[['Sol']],1,fn_compute_varpred,design_matrix=X)
herit_settle48_transform <- model$VCV[,1]/(rowSums(model$VCV)+VF)

save(settle48_transform,prior_settle48_transform,herit_settle48_transform,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle48_transform.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle48_transform.RData')
model <- settle48_transform
prior <- prior_settle48_transform
herit <- herit_settle48_transform

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
fn_plot_all_priorandpost(model,prior,xlim=0.15)
plot(herit)
```

### Binomial model

#### 24 hours
```{r,eval=FALSE, echo=FALSE}
load('data/settlement_24_binom.RData')

prior_settle24_binom = list(R=list(V=1,fix=1), #R: residual effect. Fixed at 1 for a binomial model.
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1), #G1: breeding value
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) # G2: clutch

settle24_binom <- MCMCglmm(settled~1,
                   random=~animal+clutch,
                   family = 'threshold',
                   prior=prior_settle24_binom,
                   ginverse=list(animal=Ainv_24_binom),
                   data=settlement_24_binom,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle24_binom
herit_settle24_binom <- model$VCV[,1]/rowSums(model$VCV)

save(settle24_binom,prior_settle24_binom,herit_settle24_binom,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle24_binom.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle24_binom.RData')
model <- settle24_binom
prior <- prior_settle24_binom
herit <- herit_settle24_binom

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
#fn_plot_all_priorandpost(model,prior,xlim=0.75)
plot(herit)
```

#### 48 hours

```{r,eval=FALSE, echo=FALSE}
load('data/settlement_48_binom.RData')

prior_settle48_binom = list(R=list(V=1,fix=1), #R: residual effect. Fixed at 1 for a binomial model.
                         G=list(G1=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1), #G1: breeding value
                                G2=list(V = 1,nu = 1000, alpha.mu=0, alpha.V=1))) # G2: clutch

settle48_binom <- MCMCglmm(settled~1,
                   random=~animal+clutch,
                   family = 'threshold',
                   prior=prior_settle48_binom,
                   ginverse=list(animal=Ainv_48_binom),
                   data=settlement_48_binom,nitt=1000000,burnin=10000,thin=100,verbose=TRUE)

#calculate heritability
model <- settle48_binom
herit_settle48_binom <- model$VCV[,1]/rowSums(model$VCV)

save(settle48_binom,prior_settle48_binom,herit_settle48_binom,file='C:/Users/eschlatter/Dropbox/coral_h2/analysis/models/settle48_binom.RData')
```

```{r,echo=FALSE, message=FALSE}
load('models/settle48_binom.RData')
model <- settle48_binom
prior <- prior_settle48_binom
herit <- herit_settle48_binom

summary(model)
summary(herit)
effectiveSize(herit)
plot(model)
plot(herit)
#fn_plot_all_priorandpost(model,prior,xlim=0.75)
```


## Lipids

Two samples per pair of parents (4 for the clones). Each sample has a lipids value and an egg value; we can calculate loss as (lipid-egg)/egg.

## Protein

Four samples per pair of parents (8 for the clones).